What does it take to create with domain-appropriate tools?

### 2020-01-21 convivial-2020/manuscript.tex
Polish images, rework ... ETC!

\documentclass[english,submission]{programming}
\usepackage[backend=biber]{biblatex} % Use Biblatex
\addbibresource{example.bib}

\lstdefinelanguage{JavaScript}{
  keywords={break, case, catch, continue, debugger, default, delete, do, else,
  finally, for, function, if, in, instanceof, new, return, switch, this, throw,
  try, typeof, var, void, while, with, let},
  morecomment=[l]{//},
  morecomment=[s]{/*}{*/},
  morestring=[b]',
  morestring=[b]",
  sensitive=true
}

\providecommand{\tightlist}{}% Don't want Pandoc's tight lists
  %\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

\paperdetails{
  perspective=engineering,
  area={General-purpose programming}
}

@@

\keywords{paper, showcase, lorem ipsum}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002944.10011122.10003459</concept_id>
<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}`

@@

\begin{abstract}
  This paper shows...
\end{abstract}

@@

  Surely the trick is to \emph{use} coding to build something \emph{better
  than it}. And then use that, to build something even better. But there
  is an enormous breadth and depth of philosophies here, along with many
  failed historical attempts to do better --- or at least, ones that
  failed to catch on. But even worse than this is that in my very
  \emph{language} here I am making the same mistake as the text editor ---
  speaking in unqualified terms of ``better'' and ``worse'' as if there
  really is a one-size-fits-all solution to software creation!

  Of course what we \emph{really} want is the ability for people to create
  \emph{in the way that they think is best} in their particular context
  --- to equip them to feasibly create the tools that suit them for the
  thing they want to make. And second-order tools that suit them for
  making the first-order tools, etc. It would do no good to replace
  text-imperialism with anything-else-imperialism, which is one
  interpretation of calls for alternatives. This dream goes beyond
  familiar senses of what constitutes a ``craft'', as far as a strong
  melding of tool and material.

  Certainly in the real world there is only one ``material'' --- matter
  --- and humanity as a whole does create ``higher-order'' tools. A
  (future) software craft, however, could be expected to give this power
  to \emph{individuals} instead of the community alone. The
  one-size-fits-all approach is an understandable consequence of few
  designers catering to many users --- perhaps due to the necessity to
  specialise in topics, languages, libraries, and tools. Their complexity
  is part of the general difficulty of software creation, exacerbated by
  the one-size-fits-all approach, closing the circle.

  What we need is some system that not only lets us create software in a
  way that is ``close to the problem domain'' as decided by the
  user-developer, but also can augment or change itself to adapt to a
  different ``way of creating''. Existing systems seem to only have one of
  these properties without the other. Smalltalk and LISP try to minimise
  arbitrary commitments of language \emph{semantics} to be adaptable, but
  their being textual languages is a fairly tough commitment to break out
  of. And it is not so hard to make a specific, \emph{hard-baked} visual
  or alternative programming tool --- but it is hard to make it
  re-programmable \emph{without} having to go back to \emph{its} textual
  source code.

  Again, if someone wants to type out pictures in ASCII, let them ---
  whether they do it for a challenge, or even if they find that more
  natural for themselves. But equally, if I want to do it another way,
  then please give me that affordance. This is how I segue into the
  software artefact for which I have been attempting to build a natural
  representation. True, it is a programmer's artefact, but it is still
  representative of what any normal person has to do, insofar as:

@@

  In summary, an object is a key-value mappings of state which can change
  as a result of messages received by it. A message is sent by first
  \emph{bind}-ing its name to its implementation, i.e.~specific code which
  is then run in the context of the receiver. ``Bind'' is accomplished via
  another message send, this time to the receiver's \emph{vtable}, or
  ``behaviour''. This is another object whose state defines a particular
  way of associating ``message name'' to ``implementation code''. This
  continues up the vtable-chain, and terminates at a base case. The system
  as a whole is ``bootstrapped'' into existence by its initialisation
  code. Each step of this code makes use of any parts of the system set up
  by the previous steps.

  The paper itself consists of mostly prose, several code listings, and
  full C sources for a sample implementation at the end. It also provides
  several diagrams. But to understand it, I repeatedly found myself
  drawing \emph{extra} diagrams. For example, the first acts of the
  running system boil down to initialising the three or so objects. This
  consists of allocating memory, interpreting it as a C struct and then
  filling in fields in a mundane manner. I had great difficulty following
  the specifics in my head, but when I drew tables in the style of their
  diagrams I readily saw what what going on. I personally would have
  preferred these to have been in the paper in the first place, but I am
  not necessarily representative of those who read it. The
  ``one-size-fits-all'' approach is perhaps unavoidable for static media
  like print.

@@

  (OROM/HTML screenshot)

  I was grateful for the browser's management of graphical layout,
  resizable text fields, and keeping the DOM tree synchronised with what
  you actually see. This last property enabled me to make the decision to
  \emph{directly} encode much of the system state in the DOM, achieving
  basic liveness (the thing on the screen is the actual thing) for the
  keys and values of obj-dicts. For each obj-dict I used a two-column
  \texttt{\textless{}table\textgreater{}} within a
  \texttt{\textless{}div\textgreater{}}, matching the diagrams, and gave
  the rows CSS class names matching the keys for easy lookup. Sending
  messages relied on the JavaScript console, but existing values in text
  boxes (such as method implementations) could be edited directly. Also
  thanks to the browser, I could easily make text boxes and
  \texttt{\textless{}div\textgreater{}}s resizable as necessary.

@@

  \hypertarget{orom-as-files-and-directories}{%
  \subsection{OROM as files and
  directories}\label{orom-as-files-and-directories}}

  \begin{quote}
  I would probably put this after the SVG part or perhaps even later once
  you discuss more general principles (like the idea of polyfilling to
  make a domain substrate into a programmable system). Then you can give
  this as another example.
  \end{quote}

@@

  (Screenshot of OROM/SVG showing boxes, code and arrows)

  In this version, obj-dicts are encoded as nested SVG
  \texttt{\textless{}rect\textgreater{}}s and other elements, reminiscent
  of diSessa's Boxer. This was a significant departure from the table
  representation, and even though SVG supports (some) nested HTML via
  \texttt{\textless{}foreignObject\textgreater{}}, I actually preferred
  the possibility of multiple levels of nesting.

@@

  \begin{quote}
  Why is OROM/SVG more challenging? I think it would be useful to sketch
  here, because it reveals some useful principles. There is more
  polyfilling (arrows and resizing is not in the SVG substrate itself),
  but at least it's possible to add those (unlike in HTML)
  \end{quote}

  \hypertarget{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}{%
  \section{Inevitable Requirements Of Most Software (and the work we
  must do to meet
  them)}\label{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}}

  As programmers, we have a maxim along the lines of: when you find
  yourself repeating the same thing over and over again, factor it into a
  ``thing'' and let the computer do the duplication. Some consider the
  ``design pattern'' to be simply what a ``language feature'' looks like
  when it's not part of the syntax.

  \begin{quote}
  The jump from very general abstraction idea to design patterns and
  language features in the above is not exactly clear - it would be good
  to help the reader understand why are you talking about this here.
  \end{quote}

  As I developed OROM/SVG I found myself implementing many such patterns.
  It seems that languages are often at the wrong level of abstraction for
  the requirements of modern software, necessitating the same boilerplate
  per project just to get up to basic functionality. This burden either
  falls on the author, or on the wider community to build and maintain
  higher-level frameworks, syntax extensions, etc.

  By contrast, these languages seem much better adapted to e.g.~batch-mode
  printf()-centric applications. That they fail for the \emph{common case}
  should be concerning. I will now present what seem to be the inevitable
  demands of this common case.

  \begin{quote}
  I'm not sure I understand what the above opening is trying to say - I
  suppose it might be because it is quite high-level/general. What is this
  section about more specifically? Is it about what certain kind of
  software consists of and what we need to build that? I would also be a
  bit careful about things like ``most modern software'' below. It would
  be good to instead characterize what kind of software are you interested
  in - end-user graphical systems? (a microservice or something like that
  does not need SVG\ldots{})
  \end{quote}

  \hypertarget{as-a-mere-consumer}{%
  \subsection{As A ``Mere Consumer''}\label{as-a-mere-consumer}}

  I begin by describing expectations of software that even ``end-users''
  hold, consigned as they are to more or less accept what we give them.

  \hypertarget{retained-mode-vector-graphics}{%
  \subsubsection{Retained-Mode Vector
  Graphics}\label{retained-mode-vector-graphics}}

@@

  On this requirement, SVG fits the bill very well. Although it is not
  part of the JavaScript language \emph{per se}, it is a standard and
  widely supported technology of the Web \emph{platform}. We can observe
  that anyone with a browser \emph{in principle} has access to a powerful
  vector graphics editor --- just one with no GUI. I will return to this
  later.

  The SVG tree keeps the nice properties of the DOM, such as updating the
  display when shape parameters are changed. This is well-adapted to
  ``I/O-bound'' software like mine, where things change only in response
  to user input. If I wanted animation, this boils down to a regular
  ``advance simulation'' signal, and would require setting up some
  rendering loop. Alternatively, there is the W3C's chosen ontology of CSS
  animations, but see section ??.

  \hypertarget{basic-assumptions-about-physical-objects}{%
  \subsubsection{Basic Assumptions About Physical
  Objects}\label{basic-assumptions-about-physical-objects}}

  In any software making use of vector graphics, there is usually some
  level of ``physics'' expected by users. This need not be nearly as
  exhaustive as the word ``physics'' might imply, as in e.g.~physics
  engines for games; I feel it is important to recognise it for what it is
  instead of conceiving physics as an inherently complex thing to be found
  only in specialised simulations.

  \begin{quote}
  People might not find this easy to believe - but I think you can
  convince them by saying that pretty much all mobile phone apps have some
  sort of ``phone touch physics'' (with sliding, opening menus etc.).
  \end{quote}

@@

  The level of physics in software tends to not involve force, or mass, or
  very much at all, merely position and space. One thing that all usable
  software must do, for example, is not crush many visually complex
  shapes, such as lines of text, into the same region, since it becomes
  unreadable. Such concepts of ``solid objects do not intersect'' or
  ``only things at different layers may overlap'' are basic rules
  inherited from the real world of graphical presentation.

@@

  \hypertarget{translationally-rigid-bodies}{%
  \subparagraph{Translationally Rigid
  Bodies}\label{translationally-rigid-bodies}}

@@

  \hypertarget{maintaining-relationships-over-time}{%
  \subsubsection{Maintaining Relationships Over
  Time}\label{maintaining-relationships-over-time}}

@@

  If a platform does not provide a means to causally link and unlink bits
  of live state, then this must form part of the standard boilerplate.

  Such was the case in OROM/SVG. I implemented the system in an OOP
  fashion, and the Observable class is the most widely used. It wraps a
  current value and a list of subscribers, notifying them when it changes.
  It can be seen that this supplanted JavaScript's own ``dumb'' state by
  the frequency of the change() function throughout the code.

@@

  \begin{verbatim}
  subscribe(object.position, pointer.position);
  \end{verbatim}

@@

  \hypertarget{nut-cracking-with-sledgehammers}{%
  \subsubsection{Nut-Cracking With
  Sledgehammers}\label{nut-cracking-with-sledgehammers}}

@@

  We are conditioned to only think of these in their most general forms.
  The vector graphics I use in OROM/SVG are just rects, lines, circles and
  text; a fraction of the full capability of SVG. The ``basic physics'' I
  use is dwarfed by fully general 2D or 3D physics engines. The only
  layout algorithm I had the patience to implement was a simple way to
  expand a list of boxes to fit in a new child at the bottom; the
  affordance to place and size boxes \emph{manually} staves off the rest
  for the time being! Yet search for material on layout algorithms and it
  can seem like Fully General Linear Inequality Solvers like Cassowary are
  all there is.

  The inevitable requirements I suggest here, do \emph{not} necessitate
  Fully General anything. In fact, such representations would make it more
  cumbersome to express what I wanted in OROM. As the old wisdom goes,
  there's no point expressing a simple regex search as an arbitrary Turing
  machine; my boxes don't \emph{have} a moment of inertia or a mass and I
  don't \emph{need} a linear optimisation solver for my space management
  --- for the time being.

  Unfortunately, there doesn't seem to be much interest on the
  smaller-scale instances of these problems. I suggest the name
  ``geometric physics'' for the subset of physics I have mentioned.

  \begin{quote}
  I think the above section talks about 3 things - (i) it describes some
  of the things that you had to implement for OROM/SVG, (ii) it talks
  about some general patterns that arise there (like shapes and physics)
  and (iii) it mentions some implementation details. I found it a bit
  confusing to have these all mixed together. It reads a bit as if you
  were jumping from one random topic that you encountered when
  implementing this - but that makes the section very specific to your
  implementation. How can we make it clear that there are more general
  principles here? I think it might be better if you started with some
  larger example (hypothetical is fine!) to illustrate what kind of things
  you'd like to build - e.g. imagine you're building PowerPoint in OROM -
  what would be some of the things you'd need? (for both the end
  application and the system that you're using to build it). This would
  then nicely motivate the more specific topics (graphics, physics,
  relationships).
  \end{quote}

  \hypertarget{as-a-developer}{%
  \subsection{As A ``Developer''}\label{as-a-developer}}

  I now turn to some things that were perhaps inevitable, in any case
  useful, as the \emph{implementor} of OROM/SVG.

  \hypertarget{practicality-vs.-high-aesthetics}{%
  \subsubsection{Practicality vs.~High
  Aesthetics}\label{practicality-vs.-high-aesthetics}}

  \begin{verbatim}
  Entity.prototype.addState = function(key) {
    let tr = document.createElement('tr');
    tr.setAttribute('class', key);

    let td = document.createElement('td');
    td.setAttribute('class', 'key');
    td.textContent = key;
    tr.appendChild(td);
    ...
  \end{verbatim}

  and in OROM/SVG the story is similar. It takes advantage of the powerful
  developer tools included in browsers (e.g.~the DOM inspector), and
  avoids the need to manually synchronise separate Model and View for such
  encoded state. With a little work, it also allowed the system a large
  degree of externalisability --- see Section ??. I consider this as
  practical, expedient, ``hijacking'' of these substrates because it
  either goes against their intended use (HTML tables for presentation
  rather than semantics) or are simply ``odd'' (using certain arrangements
  of shapes to ``encode'' machine-readable semantics).

  Another example of practicality was in the issue of positioning and
  sizing. In Section ?? I mentioned that I was able to sidestep some
  layout work by offloading it to the user. While building the system I
  tended to want to draw boxes into existence, and for that my brain's
  existing aesthetic algorithms can decide where they go. There are areas
  where the machine must be able to figure this out itself --- I don't
  want to be constantly bothered by prompts to place boxes as they are
  allocated by some running code --- but for one-offs it is a very
  convenient way to do a little less yak-shearing.

  \begin{quote}
  I think you could adapt the above discussion a bit and focus less on
  what you don't want (high modernism) and more on what you want - what is
  that? I think it's simplifying the problem along a different axis - to a
  certain domain with relatively simple interactions and aesthetics. (So
  rather than trying to make minimal formal calculus that's provably
  correct, you're trying to make minimal working interaction model that's
  interesting)
  \end{quote}

  \hypertarget{positioning-and-sizing}{%
  \subsubsection{Positioning and Sizing}\label{positioning-and-sizing}}

  The simple desire to move and resize boxes with the mouse motivated a
  lot of the live-state and geometric physics ideas. This problem could be
  considered a microcosm of OROM/SVG: what's a natural way I
  \emph{conceive} of this behaviour, and could I implement it that way?

  \begin{quote}
  The above point (moving motivated live state and physics ideas) would be
  a lot more obvious if you followed the structure with motivating example
  first followed by an analysis\ldots{}
  \end{quote}

  It starts with a consideration of translational rigidity. In its most
  primitive form, this is a relation between two points. Thus it is
  natural to draw a line or ``rod'' between them. It seems that this rod
  transmits changes in one of its endpoints directly to the other
  endpoint. Since they both feel the same deltas, the displacement vector
  between them is preserved.

  (Diagram of delta transmission)

@@

  Resizing (of boxes at least) could be achieved through rods that stay
  horizontal or vertical. In the language of ``small differences'' spoken
  by my live-state infrastructure, this is expressed as a rod
  ``transmitting deltas'' in the vertical and horizontal, ``absorbing''
  the other component into itself. Mirroring a DOM rect to these rods is
  as simple as subscribing its width and height to horizontal and vertical
  rods' length Observables. This way, boxes can be resized from whatever
  corner is convenient.

  (Box demonstrating all three rod types)

  I colour such ``half-rigid'' rods yellow and fully rigid rods red. The
  ``absorb all endpoint changes'' case is coloured green.

  \begin{quote}
  I don't think you explained what `half-rigid' means yet.
  \end{quote}

@@

  \begin{quote}
  This section reads more as an implementation detail - even though it
  makes some interesting points about the physics \& rods system. However,
  it might better to have it in a section that's more about implementation
  than about general principles. Or perhaps you could restructure Section
  3 to have (i) example (ii) general lessons (like physics etc.) and (iii)
  specific points about implementation (e.g.~your concrete physics system
  with rods, what you had to `polyfill')
  \end{quote}

  \hypertarget{visible-coordinate-systems}{%
  \subsubsection{Visible Coordinate
  Systems}\label{visible-coordinate-systems}}

@@

  First of all, SVG shapes (e.g.~\texttt{\textless{}rect\textgreater{}})
  are strictly leaf nodes, so if I wish to nest boxes within boxes, the
  visible ``box'' must be a mere \emph{accessory} to the nestable element,
  in my case a \texttt{\textless{}g\textgreater{}} (group). This means
  that instead of resizing the x, y, width, height attributes of the
  \texttt{\textless{}rect\textgreater{}}, only its width and height change
  along with the transform attribute of its \emph{parent}
  \texttt{\textless{}g\textgreater{}}. This was not too bad; just
  subscribe this attribute, instead of the
  \texttt{\textless{}rect\textgreater{}} position, to the top-left Point
  handle.

  All child elements of a node transform with it, so already SVG has baked
  in a basic facility for translational rigidity. This is only available
  as a tree hierarchy, but it is still useful. However, it conflicts with
  my early decision to have Point objects all share the global co-ordinate
  system. This way, connecting ``this point'' to the mouse pointer, etc.
  is not infuriating to express. Still, it was necessary in the case of
  certain elements --- especially those which must transcend the tree
  structure altogether, like arrows between boxes --- to bite this bullet,
  one way or another.

@@

  Problems: cognitive vs.~``official'' complexity

>>>>

@@

@@

\newcommand{\joel}[1]{}
\newcommand{\svgel}[1]{\texttt{\textless{}#1\textgreater{}}}
\newcommand{\OSFA}{One-Size-Fits-All}

\begin{abstract}
There is a One-Size-Fits-Allquality to languages, APIs and even programming itself. Whether you're making a mobile game or a scientific simulation, you will be using a text-based language with similar devices for structuring your code. This is a source of artificial difficulty in creating, understanding, and modifying software systems. No matter the domain, the author's design needs encoding into a form that does not resemble it.

This paper describes a vision where software can be built in a programming environment that is closer to the domain of the software itself. By doing so, users of the system can use familiar abstractions and tools for adapting it. A step towards this vision is presented: a Web version of a minimal OOP system, developed as an executable version of the diagrams of its design, in a substrate meant to facilitate this. The experience of creating such a substrate is analysed, and I suggest deficiencies in programming environments that stand in the way of making this practice commonplace, as well as ways to fill in these gaps.
\end{abstract}

\keywords{substrate, visual, context, domain, adaptation}

@@

Surely the trick is to \emph{use} coding to build something \emph{better
than it}. And then use that, to build something even better. But there
is an enormous breadth and depth of philosophies here, along with many
failed historical attempts to do better --- or at least, ones that
failed to catch on. But even worse than this is that in my very
\emph{language} here I am making the same mistake as the text editor ---
speaking in unqualified terms of ``better'' and ``worse'' as if there
really is a One-Size-Fits-Allsolution to software creation!

Of course what we \emph{really} want is the ability for people to create
\emph{in the way that they think is best} in their particular context
--- to equip them to feasibly create the tools that suit them for the
thing they want to make. And second-order tools that suit them for
making the first-order tools, etc. It would do no good to replace
text-imperialism with anything-else-imperialism, which is one
interpretation of calls for alternatives.

This dream goes beyond the familiar sense of what constitutes a
``craft'', as far as a strong melding of tool and material. Parallels
can be drawn with industrialisation and a strong division of labour: the
community as a whole produces its higher-order tools, but currently no
single person can have the same autonomy. A (future) software craft
could be expected to give this power to \emph{individuals}, instead of
the community alone. Whenever there are many small specialities
(e.g.~languages, tools, or subject areas) each serving many clients, the
One-Size-Fits-Allstyle is the best one can hope for. Adaptation to
individual preferences and idiosyncrasies is only feasible when those
individuals can do it themselves.

What we need is some system that not only lets us create software in a
way that is ``close to the problem domain'' as decided by the
user-developer, but also can augment or change itself to adapt to a
different ``way of creating''. Existing systems seem to only have one of
these properties without the other: Smalltalk and LISP try to minimise
arbitrary commitments of language \emph{semantics} to this end, but
their being textual languages is a fairly tough commitment to break out
of. And it is not so hard to make a specific, \emph{hard-baked} visual
or alternative programming tool --- but it is hard to make it
re-programmable \emph{without} having to go back to \emph{its} textual
source code.

If someone wants to type out pictures in ASCII, let them --- whether
they do it for a challenge, or even if they find that more natural for
themselves. But equally, if I want to do it another way, then please
give me that affordance. This is how I segue into the software artefact
for which I have been attempting to build a natural representation.
True, it is a programmer's artefact, but it is still representative of
what any normal person has to do, insofar as:

@@

Essentially, an OROM object is a block of state which can change as a
result of messages received by it. A message is sent by first
\emph{bind}-ing its name to its \emph{implementation}: specific code,
which is then run in the context of the receiver. ``Bind'' is
accomplished via another message send, this time to the receiver's
\emph{vtable}, or ``behaviour''. This is another object whose state
defines a particular way of associating ``message name'' to
``implementation code''. This continues up the vtable-chain, and
terminates at a base case. The system as a whole is ``bootstrapped''
into existence by its initialisation code. Each step of this code makes
use of any parts of the system set up by the previous steps.

The paper itself consists of mostly prose, several code listings, and
full C sources for a sample implementation at the end. It also provides
several diagrams. But to understand it, I repeatedly found myself
drawing \emph{extra} diagrams. For example, the first acts of the
running system boil down to initialising the three or so objects. This
consists of allocating memory, interpreting it as a C struct and then
filling in fields in a mundane manner. I had great difficulty following
the specifics in my head, but when I drew tables in the style of their
diagrams I readily saw what what going on. I personally would have
preferred these to have been in the paper in the first place, but I am
not necessarily representative of those who read it. The
One-Size-Fits-Allapproach is perhaps unavoidable for static media like
print.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-html.png}
  \caption{OROM/HTML: Obj-dicts are rendered as resizable tables, which reference
           each other through numerical IDs. \label{fig:orom-html}}
\end{figure}

To emphasise the tendency of OROM objects to be visualised as key-value
mappings, I will refer to them as \emph{obj-dicts}. Figure
\ref{fig:orom-html} shows the OROM/HTML implementation, in which they
take the form of HTML tables.

Here, I was grateful for the browser's management of graphical layout,
resizable text fields, and keeping the DOM tree synchronised with what
one sees. This last property enabled me to make the decision to
\emph{directly} encode much of the system state in the DOM, achieving
basic liveness\footnote{``The thing on the screen is the actual thing''}
for the keys and values of obj-dicts.

For each obj-dict I used a two-column
\texttt{\textless{}table\textgreater{}} within a
\texttt{\textless{}div\textgreater{}}, matching the diagrams, and gave
the rows CSS class names matching the keys for easy lookup. Sending
messages relied on the JavaScript console, but existing values in text
boxes (such as method implementations) could be edited directly. Also
thanks to the browser, I could easily make text boxes and
\texttt{\textless{}div\textgreater{}}s resizable as necessary.

@@

\hypertarget{practicality-vs.-high-aesthetics}{%
\subsection{Practicality vs.~High
Aesthetics}\label{practicality-vs.-high-aesthetics}}

\begin{lstlisting}[language=JavaScript]
Entity.prototype.addState = function(key) {
  let tr = document.createElement('tr');
  tr.setAttribute('class', key);

  let td = document.createElement('td');
  td.setAttribute('class', 'key');
  td.textContent = key;
  tr.appendChild(td);
  ...
\end{lstlisting}

and in OROM/SVG the story is similar. It takes advantage of the powerful
developer tools included in browsers (e.g.~the DOM inspector), and
avoids the need to manually synchronise separate Model and View for such
encoded state. With a little work, it also allowed the system to be
saved and restored --- see Section \ref{persistence}. I consider this as
practical, expedient, ``hijacking'' of these substrates because it
either goes against their intended use (HTML tables for presentation
rather than semantics) or are simply ``odd'' (using certain arrangements
of shapes to ``encode'' machine-readable semantics).

Another example of practicality was in the issue of positioning and
sizing. In Section \ref{nut-cracking-with-sledgehammers} I mentioned
that I was able to sidestep some layout work by offloading it to the
user. While building the system I tended to want to draw boxes into
existence, and for that my brain's existing aesthetic algorithms can
decide where they go. There are areas where the machine must be able to
figure this out itself --- I don't want to be constantly bothered by
prompts to place boxes as they are allocated by some running code ---
but for one-offs it is a very convenient way to do a little less
yak-shearing.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../orom-svg.png}
  \caption{OROM/SVG: Obj-dicts are moveable and resizable nested boxes,
           referencing each other via real arrows.\label{fig:orom-svg}}
\end{figure}

In this version, shown in Figure \ref{fig:orom-svg}, obj-dicts are
encoded as nested SVG \texttt{\textless{}rect\textgreater{}}s and other
elements, reminiscent of diSessa's Boxer. This was a significant
departure from the table representation, and even though SVG supports
(some) nested HTML via \texttt{\textless{}foreignObject\textgreater{}},
I actually preferred the possibility of multiple levels of nesting.

@@

\hypertarget{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}{%
\section{Typical Requirements Of Common Software (and the work we must
do to meet
them)}\label{typical-requirements-of-common-software-and-the-work-we-must-do-to-meet-them}}

The ``common case'' of software present in our daily lives shares
certain properties, such as being graphical and interactive. Such are
expectations that ``end-users'' hold, consigned as they are to merely
\emph{consume} what programmers give them. But if we want to make
``programming'' more like ``using'', such expectations on \emph{use} at
least need to be acknowledged. In this section, I will present these and
other seemingly \emph{inevitable} demands of normal software,
exemplified by OROM/SVG. I will comment how our programming platforms
measure up to the task of helping us meet them, including my chosen
platform of JavaScript and Web technologies.

\hypertarget{retained-mode-vector-graphics}{%
\subsection{Retained-Mode Vector
Graphics}\label{retained-mode-vector-graphics}}

@@

On this requirement, SVG fits the bill very well. Although it is not
part of the JavaScript language \emph{per se}, it is a standard and
widely supported technology of the Web \emph{platform}. We can observe
that anyone with a browser \emph{in principle} has access to a powerful
vector graphics editor --- just one with no GUI. I will return to this
at the end (Section ??).

The SVG tree keeps the nice properties of the DOM, such as updating the
display when shape parameters are changed. This is well-adapted to
``I/O-bound'' software like mine, where things change only in response
to user input. If I wanted animation, this boils down to a regular
``advance simulation'' signal, and would require setting up some
rendering loop. Alternatively, there is the W3C's chosen ontology of CSS
animations, but see Section \ref{context-appropriate-ontologies}.

\hypertarget{basic-assumptions-about-physical-objects}{%
\subsection{Basic Assumptions About Physical
Objects}\label{basic-assumptions-about-physical-objects}}

In any software making use of vector graphics, there is usually some
level of ``physics'' expected by users. This need not be nearly as
exhaustive as the word ``physics'' might imply, as in e.g.~physics
engines for games; I feel it is important to recognise it for what it is
instead of conceiving physics as an inherently complex thing to be found
only in specialised simulations. For example, pretty much all mobile
apps have what could be called ``phone touch physics'' where menus and
screens slide in and out.

@@

The level of physics in software tends to not involve force, or mass, or
very much at all, merely position and space; we could call it
``geometric physics''. One thing that all usable software must do, for
example, is not crush many visually complex shapes, such as lines of
text, into the same region, since it becomes unreadable. Such concepts
of ``solid objects do not intersect'' or ``only things at different
layers may overlap'' are basic rules inherited from the real world of
graphical presentation.

@@

\hypertarget{translationally-rigid-bodies}{%
\subsubsection{Translationally Rigid
Bodies}\label{translationally-rigid-bodies}}

@@

\hypertarget{maintaining-relationships-over-time}{%
\subsection{Maintaining Relationships Over
Time}\label{maintaining-relationships-over-time}}

@@

If a platform does not provide a means to causally link and unlink bits
of live state, then this must form part of the standard boilerplate.
Such was the case in OROM/SVG. I implemented the system in an OOP
fashion, and the Observable class is the most widely used. It wraps a
current value and a list of subscribers, notifying them when it changes.

@@

\begin{lstlisting}[language=JavaScript, numbers=none]
subscribe(object.position, pointer.position);
\end{lstlisting}

@@

\hypertarget{nut-cracking-with-sledgehammers}{%
\subsection{Nut-Cracking With
Sledgehammers}\label{nut-cracking-with-sledgehammers}}

@@

We are conditioned to only think of these in their most general forms.
The vector graphics I use in OROM/SVG are just rects, lines, circles and
text; a fraction of the full capability of SVG. The ``geometric
physics'' I use is dwarfed by fully general 2D or 3D physics engines.
The only layout algorithm I had the patience to implement was a simple
way to expand a list of boxes to fit in a new child at the bottom; the
affordance to place and size boxes \emph{manually} staves off the rest
for the time being! Yet search for material on layout algorithms and it
can seem like Fully General Linear Inequality Solvers like Cassowary are
all there is.

The inevitable requirements I suggest here, do \emph{not} necessitate
Fully General anything. In fact, such representations would make it
\emph{more} cumbersome to express what I wanted in OROM. As the old
wisdom goes, there's no point expressing a simple regex search as an
arbitrary Turing machine; my boxes don't \emph{have} a moment of inertia
or a mass and I don't \emph{need} a linear optimisation solver for my
space management --- for the time being. Under the theme of
domain-appropriate tools, it is worth designing interfaces for
smaller-scale instances of these areas and exploring their capabilities.

\hypertarget{patterns-and-polyfilling}{%
\section{Patterns and Polyfilling}\label{patterns-and-polyfilling}}

The message of the previous section is that existing platforms are often
at the wrong level of abstraction for the requirements of common
software. I recognise that they are reasonably well-adapted to batch
mode file I/O tasks, but that they fail for the \emph{common case} is a
problem. It necessitates largely the same ``boilerplate'' setup per
project just to get basic functionality:

\begin{itemize}
\tightlist
\item
  Here is how I shall describe shapes
\item
  Here is how I make these shapes move together
\item
  Here is how I maintain internal consistency as the user unpredictably
  changes things
\end{itemize}

This burden either falls on the author, or on the wider community to
build and maintain higher-level frameworks, syntax extensions, etc. In
this paper, I refer to this process as ``polyfilling'', with an emphasis
on the DIY, individual level.

It is true, we already have a term for bringing into existence some
software feature that isn't already there: programming. The difference
is that polyfilling is about filling in \emph{boilerplate},
i.e.~functionality \emph{that should have been there in the first place,
but isn't}. This injects some subjectivity and value judgement into the
term. But ultimately, all platforms are designed with certain features
``out of the box'' and leave other features to be implemented ``if
needed''. And when a problem domain ``requires batteries'' yet the tools
available say ``batteries not included'', it is quite reasonable to want
the batteries to be part of the platform.

This section covers the boilerplate I found myself polyfilling in
OROM/SVG.

Is to bring things into the mindset / ontology of the specific software.
For example, wrapping the official Web conventions for event handling in
the Observable infrastructure, or wrapping DOM nodes with ``user data''
objects. The latter case, takes advantage of JavaScript's ability to
simply ``add on'' new fields to objects, even objects of Web APIs, to
``annotate'' the DOM node with a JavaScript helper. Very advantageous to
permit this in JavaScript. In another language like C, such would
require saving the address in a lookup table elsewhere.

\hypertarget{positioning-and-sizing}{%
\subsection{Positioning and Sizing}\label{positioning-and-sizing}}

The simple desire to move and resize boxes with the mouse motivated a
lot of the concepts in Sections
\ref{basic-assumptions-about-physical-objects} and
\ref{maintaining-relationships-over-time}. This problem could be
considered a microcosm of OROM/SVG: what's a natural way I
\emph{conceive} of this behaviour, and could I implement it that way?

It starts with a consideration of translational rigidity. In its most
primitive form, this is a relation between two points. Thus it is
natural to draw a line or ``rod'' between them. It seems that this rod
transmits changes in one of its endpoints directly to the other
endpoint. Since they both feel the same deltas, the displacement vector
between them is preserved (Figure \ref{fig:delta-transmission}).

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../delta-transmission.png}
  \caption{Fully rigid rods transmit changes in one end to the other end. \label{fig:delta-transmission}}
\end{figure}

@@

Resizing of boxes could be achieved through rods that stay horizontal or
vertical. In the language of ``small differences'' spoken by the
live-state infrastructure, this is expressed as a rod ``transmitting
deltas'' in the vertical and horizontal, ``absorbing'' the other
component into itself (Figure \ref{fig:rods}). Mirroring a DOM rect to
these rods is as simple as subscribing its width and height to
horizontal and vertical rods' \texttt{length} Observables. This way,
boxes can be resized from whatever corner is convenient.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.75\textwidth]{../rods.png}
  \caption{The yellow rods on the box border are ``half-rigid'': they transmit horizontal or vertical changes, but not both. The green rod in the top-left absorbs all changes from an endpoint into itself, visualising the displacement vector of the child box from its parent. The remaining red rods are fully rigid. A line of CSS can show this ``scaffolding'' which is normally hidden for users. \label{fig:rods}}
\end{figure}

@@

\hypertarget{visible-coordinate-systems}{%
\subsection{Visible Coordinate
Systems}\label{visible-coordinate-systems}}

@@

First of all, SVG shapes (e.g.~\texttt{\textless{}rect\textgreater{}})
are strictly \emph{leaf} nodes of the DOM. So if I wish to nest boxes
within boxes, the visible box \texttt{\textless{}rect\textgreater{}}
must be a mere \emph{accessory} to the nestable element, in my case a
\texttt{\textless{}g\textgreater{}} (group). This means that instead of
resizing the \texttt{x}, \texttt{y}, \texttt{width}, \texttt{height}
attributes of the \texttt{\textless{}rect\textgreater{}}, only its
\texttt{width} and \texttt{height} change, along with the
\texttt{transform} attribute of its \emph{parent}
\texttt{\textless{}g\textgreater{}}. This was not too bad; just
subscribe this attribute, instead of the
\texttt{\textless{}rect\textgreater{}} position, to the top-left Point
handle.

All child elements of a node transform with it, so already SVG has baked
in a basic facility for translational rigidity. This is only available
as a tree hierarchy, but it is still useful. However, it conflicts with
my early decision to have Point objects all share the global co-ordinate
system (this was to ensure that simple relations, such as a point
following the mouse pointer, are not infuriating to express.) Still, it
was necessary in the case of certain elements --- especially those which
must transcend the tree structure altogether, like arrows between boxes
--- to bite this bullet, one way or another.

@@
